$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in the Presence of Non-Functional Properties

### 2013-12-30 JournalToSubmit2013/intro.tex
another version ...still  needs examples, use case finalization, and paper organization

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
combined to produce applications and provide a business logic.
The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually being developed.

Functional properties are not the only  aspect in the software development process.
Other non-functional properties need to be addressed to fit in the application with its context (e.g., data privacy, exception handling, atomicity, data persistence).
Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
Most methods integrate them only after the application has been implemented.
This leads to service based applications that are partly specified and, thereby, partialy compliant with the requirements of the application.

@@

Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse. Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods. Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols. Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented. In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.

@@

The literature stresses the need for methodologies and techniques for service oriented analysis and design, since they are the cornerstone in the development of meaningful service based applications~\cite{5}. Existing approaches argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}. Model Driven Development (MDD)  for the development of software systems mainly characterised by the use of models as a product \cite{Selic03}.
The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}. MDA states a set of guidelines for structuring the specifications of software systems. The technique promotes  the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}).
These models are successively refined from abstract specifications into actual computer programs.

This paper proposes  extension of the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
%~\cite{decastro1}, 
to support non-functional aspects of service-oriented applications, from the early stages of software development.
$\pi$-SODM is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications. It is intended:
%SOD-M does not provide support for the specification of non-functional requirements, such as
%security, reliability, and efficiency.
\begin{trivlist}
\item \textit{(i)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
\item \textit{(ii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
\end{trivlist}

The main contribution of our work is a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements. The methodology is based on  an NFR model including a set of concepts need for the modeling of NFR in service-oriented applications.

Accordingly, the remainder of the paper is organized as follows.
Section\dots

>>>>

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
combined to build applications that provide a business logic.
The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually being developed.

Functional properties are not the only  aspect in the software development process.
Other non-functional properties need to be addressed to fit in the application with its context (e.g., data privacy, exception handling, atomicity, data persistence).
Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
Most methods integrate them only after the application has been implemented.
This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.

@@

Service oriented computing promotes ease of information systems construction thanks to services reuse. Yet, this is not applied to non-functional constraints because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods. Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols. Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented. In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.

@@

The literature stresses the need for methodologies and techniques for service oriented analysis and design, since they are the cornerstone in the development of meaningful service based applications~\cite{5}. Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software that serves, rather than dictates the needs of its users~\cite{watson}. Model Driven Development (MDD)  for the development of software systems mainly characterised by the use of models as a product \cite{Selic03}.
%The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}. 
Model Driven Architecture (MDA)~\cite{miller} states a set of guidelines for structuring the specifications of software systems. The technique promotes  the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}).
These models are successively refined from abstract specifications into actual computer programs.

This paper proposes  extension of the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
%~\cite{decastro1}, 
to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
$\pi$-SODM is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications. It is intended:
%SOD-M does not provide support for the specification of non-functional requirements, such as
%security, reliability, and efficiency.
\begin{trivlist}
\item \textit{(i)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
\item \textit{(ii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
\end{trivlist}

%The main contribution of our work is a methodology for supporting the construction of service-oriented applications,

 The methodology is based on  an NFR model including a set of concepts  for  modeling  NFR in service-oriented applications.
Accordingly, the remainder of the paper is organized as follows.
Section\dots

### 2013-12-30 JournalToSubmit2013/metamodels.tex
another version ...still  needs examples, use case finalization, and paper organization

The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model classes.} and the {\sc Actions} they perform. 
Instances of this meta-model are represented as UML activity diagrams. 
In Figure~\ref{fig:e-scomposition-metamodel} 
 Coloured boxes illustrate classes  modelling  non-functional properties and 
 white boxes represent classes modelling functional ones.

%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions. 
They are graphically presented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as an activity. 
Each action represents some type of transformation or processing. 
There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
\begin{figure}[t]
\centering
\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
\label{fig:e-scomposition-metamodel}
\end{figure}

\item The {\sc ServiceActivity} element is a composite activity that must be carried out as part of a business service and is composed by one or more executable nodes.

\item In order to represent constraints associated to services compositions, we extended introduced the concepts {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules. 
Intuitively, the conditions of each rule will be checked.
In case of no compliance, the actions defined by the rule will be performed.
The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}

@@

PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as B	PEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic. 
Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
\begin{itemizedTrivlist}
\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
A {\sc Compound Operation} is defined using an {\sc Operator}.
The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.

\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
A brie description of these classes is given next.
\end{itemizedTrivlist}
%
\begin{figure}[t]
\centering
\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Metamodel}
\label{fig:metamodel}
\end{figure}

>>>>

The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
Instances of this meta-model are represented as UML activity diagrams. 
In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
 white boxes represent classes modelling functional ones.

%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
An instance of this meta-class is graphically represented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
A class action represents some type of transformation or processing. 
There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
\begin{figure}[t]
\centering
\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
\label{fig:e-scomposition-metamodel}
\end{figure}

\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.

\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules. 
Intuitively, the conditions of each rule will be checked.
In case of no compliance, the actions defined by the rule will be performed.
The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}

@@

PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic. 
Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
\begin{itemizedTrivlist}
\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
A {\sc Compound Operation} is defined using an {\sc Operator}.
The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.

\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
A brief description of these classes is given next.
\end{itemizedTrivlist}
%
\begin{figure}[t]
\centering
\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Metamodel}
\label{fig:metamodel}
\end{figure}

### 2013-12-30 JournalToSubmit2013/modelingFlyingPig.tex
another version ...still  needs examples, use case finalization, and paper organization

$\Pi$-SODM uses two models at the CIM level (see Section~\ref{sec:modelingWithPISODM}): The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
The former is a simple model just to identify the transference of value information between components of the system.
The BPMN model establishes which are the actors and main tasks of the application.

@@

\begin{itemize}
\item {\color{magenta} Placido and Umberto: Could you please complete these items? --G \& M.}
\end{itemize}

@@

>>>>

$\Pi$-SODM uses two models at the CIM level: The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
The former  identifies the transference of value information between components of the system.
The BPMN model establishes which are the actors and main tasks of the application.

@@

@@

As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way. Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse. In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.

### 2013-12-30 JournalToSubmit2013/modelingWithPISODM.tex
another version ...still  needs examples, use case finalization, and paper organization

$\pi$-SODM meta-models represent both the functional aspects of the application as well as its non-functional constraints.
As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM meta-models are organized into three levels: 
\begin{itemize}
\item CIM (\textit{Computational Independent Models}). This level focusses on the
environment of the system, as well as on its business and requirement specifications.
This viewpoint represents the software system at its highest level of abstraction.
At this stage of the development, the structure and system processing details are still unknown or undetermined.  $\pi$-SODM defines the \textit{value}
and \textit{BPMN} meta-moels for this purpose. 
%Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) 
%and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).

@@

 The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.Non-functional requirements are defined as \textit{constraints} over actions and data.

@@

Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system. 
The following sections describe the $\pi$-SODM meta-models.

>>>>

%$\pi$-SODM meta-models represent both the functional aspects of the application as well as its non-functional constraints.
As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM proposes meta-models  organized into three levels: 
\begin{itemize}
\item CIM (\textit{Computational Independent Models}). This level focusses on the
environment of the system, as well as on its business and requirement specifications.
This level represents the software system at its highest level of abstraction.
At this stage of the development, the structure and system processing details are still unknown or undetermined.  $\pi$-SODM defines the \textit{value}
and \textit{BPMN} meta-moels for this purpose. 
%Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) 
%and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).

@@

 The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements. Non-functional requirements are defined as \textit{constraints} over actions and data.

@@

Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMs); next,  starting from such models build platform independent models (PIMs) specifying the system details; next transform such models into platform specific models (PSMs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system. 
The following sections describe the $\pi$-SODM meta-models.
